Detecting Stylistic Variation in Pop Production

### 2023-11-30 outputs/paper/paper.qmd
intro reshaping

---
title: "Detecting Stylistic Variation in Pop Production"
author: 
  - name: Michaela Drouillard
    email: "michaela.drouillard@mail.utoronto.ca"
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop. We thank Rohan Alexander and Inessa De Angelis for valuable suggestions."
date: "November 24 2023"
abstract: "We explore whether we can predict if a song was produced by Jack Antonoff using the audio features provided by Spotify's API. Our logistic regression model has an accuracy of 76%, a precision of 39%, and a recall of 76%. Our random forest model has an accuracy of 61%, and a No Information Rate of 43%. We found that danceability, a metric trained on human-tagged data, was a key predictor. Our results have implications for computational explorations of culture, and understanding how the combination of human subjectivity and machine listening can come to create an approximate for style."
format: pdf
number-sections: true
bibliography: references.bib
csl: modern-language-association.csl
---

@@

Style has been extensively studied in the field of computational stylistics, with researchers applying methods ranging from (x, y, z) to define and attribute aesthetics and patterns to artists.

We are interested in understanding how to create an approximation for production style using a data set built using the Spotify audio features API. To do this, we explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We find that, in both models, `danceability` is the strongest predictor of whether a song has been produced by Antonoff. Additionally, when we look at how danceability relates to other features in the dataset, we find that the differences in these relationships are very informative. Specifically, the degree to which danceability's relationships with other features vary helps us most accurately determine if Antonoff produced a song. Put simply, using danceability along with its interactions with other dataset features gives us the most accurate predictions.

In an interview with Spotify principal engineer Glen McDonald, we learned that the danceability and valence features were built by Spotify using crowdsourced methods.

Our results suggest that we can capture something about Antonoff's music with these variables. It might, in the end, just be that subjective response that some people feel, and others enjoy, and that we are all just trying to describe in using different methods and tools.

With the understanding that what we understand as "data-driven" in recommendation systems is often intertwined with human subjectivity -- engineers' understandings of taste and taste-making shape the recommendation systems that they build (@citeSeaver) -- we contribute to a more informed understanding of how style and tastes can been algorithmically rendered using Spotify's audio features.

>>>>

---
title: "Detecting Stylistic Variation in Pop Production"
author: 
  - name: Michaela Drouillard
    email: "michaela.drouillard@mail.utoronto.ca"
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop. We thank Rohan Alexander, Carl Wilson and Inessa De Angelis for valuable suggestions."
date: "November 24 2023"
abstract: "This study investigates the feasibility of predicting whether a song is produced by Jack Antonoff using Spotify's audio feature API. We developed two models: a logistic regression model and a random forest model. The logistic regression model, focusing on Antonoff's collaborative works, demonstrates an accuracy of 76%, precision of 39%, and a recall of 76%. The random forest model, which includes discographies of other notable pop producers, shows an accuracy of 61% and compares favorably to a No Information Rate of 43%. A significant finding is that 'danceability', a metric developed from human-tagged data, emerges as a crucial predictor in both models. This research contributes to the broader understanding of computational analysis in cultural contexts, illustrating how a blend of human judgment and algorithmic processing can approximate musical styles"
format: pdf
number-sections: true
bibliography: references.bib
csl: modern-language-association.csl
---

@@

We are interested in understanding how to create an approximation for production style based on a data set built using the Spotify audio features API. To do this, we explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We find that, in both models, `danceability` is the strongest predictor of whether a song has been produced by Antonoff. Additionally, when we look at how danceability relates to other features in the dataset, we find that the differences in these relationships are informative: the degree to which danceability's relationships with other features vary helps us most accurately determine if Antonoff produced a song. In other words, using danceability along with its interactions with other dataset features gives us the most accurate predictions.

In an interview with Spotify principal engineer Glen McDonald, we learned that the danceability and valence features were built using crowdsourced methods. With the understanding that what we understand as "data-driven" in recommendation systems is often intertwined with human subjectivity -- engineers' understandings of taste and taste-making also shape the recommendation systems that they build (@citeSeaver) -- our results suggest that it's possible to create representations of style by combining features derived from both automated audio analysis and more crowdsourced, subjective contributions. This research adds to a more informed understanding of how style and tastes can been algorithmically rendered using Spotify's audio features.
